*F1000Research* Software Tool Article Template

### 2017-09-15 paper/paper_BiocF1000.Rmd

---
title: "*F1000Research* Software Tool Article Template"
author: 
  - name: Nima Hejazi
    affiliation: Address of author1
  - name: Alan E. Hubbard
    affiliation: Address of author2
  - name: Mark J. van der Laan
    affiliation: Address of author2
abstract: Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized.
keywords: Please list up to eight keywords to help readers interested in your article find it more easily.
#bibliography: paper_BiocF1000.bib
output: BiocWorkflowTools::f1000_article
---

@@

>>>>

---
title: "*F1000Research* Software Tool Article Template"
author:
  - name: Nima S. Hejazi
    affiliation: Address of author1
  - name: Rachael Phillips
    affiliation: Address of author1
  - name: Alan E. Hubbard
    affiliation: Address of author2
  - name: Mark J. van der Laan
    affiliation: Address of author2
abstract: Abstracts should be up to 300 words and provide a succinct summary of
  the article. Although the abstract should explain why the article might be
  interesting, care should be taken not to inappropriately over-emphasise the
  importance of the work described in the article. Citations should not be used
  in the abstract, and the use of abbreviations should be minimized.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, variable importance, targeted minimum loss-based estimation.
bibliography: paper_BiocF1000.bib
output: BiocWorkflowTools::f1000_article
---

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in controlling gene expression. It remains perhaps the best studied
biological mechanism amongst all known epigenetic processes of its kind. With
relatively recent advances in computational biology and bioinformatics, probing
DNA methylation has become rather easily possible -- in fact, modern assays
allow for DNA methylation signatures at up to $850,000$ CpG sites to be measured
simultaneously. Most statistical techniques available for the analysis of data
produced by such assays rely on (generalized) linear models. Here, we provide an
alternative to such approaches. Specifically, we provide a range of _variable
importance measures_ (__VIM__), parameters that arise in statistical causal
inference, for which targeted minimum loss-based estimates may be readily
computed based on the data made available by DNA methylation assays. `methyvim`
is an R package that provides facilities for performing differential methylation
analyses within exactly this scope.

For a general discussion of the framework of targeted minimum loss-based
estimation and the role this approach plays in statistical causal inference, the
interested reader is invited to consult @vdl2011targeted and @vdl2017targeted.
For a more general introduction to (statistical) causal inference,
@pearl2009causality and @hernan2018causal may be of interest.

---

## Methodology

The core functionality of this package is made available via the eponymous
`methyvim` function, which implements a statistical algorithm designed to
compute targeted estimates of VIMs, defined in such a way that the VIMs
represent parameters of scientific interest in computational biology
experiments; moreover, these VIMs are defined such that they may be estimated in
a manner that is very nearly assumption-free, that is, within a fully
nonparametric statistical model. __The statistical algorithm consists in several
major steps:__

1. Pre-screening of genomic sites is used to isolate a subset of sites for
  which there is cursory evidence of differential methylation. For the sake of
  computational feasibility, targeted minimum loss-based estimates of VIMs are
  computed only for this subset of sites. Currently, the available screening
  approaches adapts core routines from the
  [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
  package, thought future releases will support functionality from other
  packages (e.g., [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)).

2. Nonparametric estimates of VIMs (for the specified target parameter) are
  currently computed by adapting routines from the
  [`tmle`](https://CRAN.R-project.org/package=tmle) R package. Future releases
  will support doubly-robust estimates of these VIMs (via the
  [`drtmle`](https://cran.r-project.org/web/packages/drtmle/index.html) package)
  and add parameters for continuous treatments/exposures (via the
  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).

3. Since pre-screening is performed prior to estimating VIMs, we make use of a
  multiple testing correction uniquely suited to such settings. Due to the
  multiple testing nature of the estimation problem, a variant of the Benjamini
  & Hochberg procedure for controlling the False Discovery Rate (FDR) is applied
  [@benjamini1995controlling]. Specifically, we apply the modified marginal
  Benjamini & Hochberg step-up False Discovery Rate controlling procedure for
  multi-stage analyses (FDR-MSA), which is guaranteed to control the FDR as if
  all sites were tested [@tuglus2009modified].

---

## Parameters of Interest

For discrete-valued treatments or exposures:

* The _average treatment effect_ (ATE): The effect of a binary exposure or
  treatment on the observed methylation at a target CpG site is estimated,
  controlling for the observed methylation at all other CpG sites in the same
  neighborhood as the target site, based on an additive form. In particular, the
  parameter estimate represents the __additive difference__ in methylation that
  would have been observed at the target site had all observations received the
  treatment versus the scenario in which none received the treatment.

* The _relative risk_ (RR): The effect of a binary exposure or treatment on the
  observed methylation at a target CpG site is estimated, controlling for the
  observed methylation at all other CpG sites in the same neighborhood as the
  target site, based on an geometric form. In particular, the parameter estimate
  represents the __multiplicative difference__ in methylation that would have
  been observed at the target site had all observations received the treatment
  versus the scenario in which none received the treatment.

Support for continuous-valued treatments or exposures is _planned but not yet
available_, though work is underway to incorporate into our methodology the
following

* A _nonparametric variable importance measure_ (NPVI) [@chambaz2012estimation]:
  The effect of continous-valued exposure or treatment (the observed methylation
  at a target CpG site) on an outcome of interest is estimated, controlling for
  the observed methylation at all other CpG sites in the same neighborhood as
  the target (treatment) site, based on a parameter that compares values of the
  treatment against a reference value taken to be the null. In particular, the
  implementation provided is designed to assess the effect of differential
  methylation at the target CpG site on a (typically) phenotype-level outcome
  of interest (e.g., survival), in effect providing an nonparametric evaluation
  of the impact of methylation at the target site on said outcome.

_As previously noted, in all cases, an estimator of the target parameter is
constructed via targeted minimum loss-based estimation._

Having now discussed the foundational principles of the estimation procedure
employed and the statistical algorithm implemented, it is best to proceed by
examining `methyvim` by example.

---

## Preliminaries: Setting up the Data

```{r reqs, echo=FALSE}
suppressMessages(library(tmle))
suppressMessages(library(minfi))
suppressMessages(library(SummarizedExperiment))
```

First, we'll load the `methyvim` package and the example data contained in the
`methyvimData` package that accompanies it:

```{r prelims}
library(methyvim)
library(methyvimData)
```

Now, let's load the data set and seed the RNG:

```{r get-data}
data(grsExample)
set.seed(479253)
grsExample
```

The example data object is of class `GenomicRatioSet`, provided by the `minfi`
package. The summary provided by the `print` method gives a wealth of
information on the experiment that generated the data -- since we are working
with a simulated data set, we need not concern ourselves with much of this
information. While most slots of objects of this S4 class have descriptive
names, the interested analyst might consider consulting the documentation of the
`minfi` package for an in-depth description [@aryee2014minfi].

---

## Quantifying the Effect of an Exposure on DNA Methylation

### The Average Treatment Effect as Variable Importance Measure

...

```{r methyvim-ate}
suppressMessages(
  methyvim_out_ate <- methyvim(data_grs = grsExample, sites_comp = 10,
                               var_int = 1, vim = "ate", type = "Mval",
                               filter = "limma", filter_cutoff = 0.05,
                               parallel = FALSE, tmle_type = "sl"
                              )
)
methyvim_out_ate
```

As is clear from examining the object `methyvim_out_ate`, the output resembles
exactly that returned when examining objects of class `GenomicRatioSet` from the
`minfi` R package. In particular, the returned `methytmle` object is merely a
modified form (in particular, a subclass) of the input `GenomicRatioSet` object
-- thus, it contains all of the original slots, with all experimental data
intact. Several extra pieces of information are contained within the output
object as well_.

We can take a look at the results produced from the estimation procedure by
examining the `"vim"` `slot` of the `methytmle` object:

```{r methyvim-ate-print}
head(slot(methyvim_out_ate, "vim"))
```

From the table displayed, we note that we have access to point estimates of the
ATE ("est_ATE") as well as lower and upper confidence interval bounds for each
estimate ("lowerCI_ATE" and "upperCI_ATE", respectively). Additional statistical
information we have access to include the variance ("Var_ATE") of the estimate
as well as the p-value ("pval") associated with each estimate (based on
Wald-style testing procedures). Beyond these, key bioinformatical quantities
(with respect to the algorithm outlined above) are also returned; these include
the total number of neighbors of the target site, the number of neighboring
sites controlled for when estimating the effect of exposure on DNA methylation,
and, finally, the maximum correlation between the target site and any given site
in its full set of neighbors.

...

---

### The Risk Ratio as Variable Importance Measure

...

```{r methyvim-rr}
methyvim_out_rr <- methyvim(data_grs = grsExample, sites_comp = 10,
                            var_int = 1, vim = "rr", type = "Mval",
                            filter = "limma", filter_cutoff = 0.05,
                            parallel = FALSE, tmle_type = "sl"
                           )
methyvim_out_rr
```
...

```{r methyvim-rr-print}
head(slot(methyvim_out_rr, "vim"))
```
...

---

## Visualization of Results

...

```{r methyvim-plot-volcano, eval=FALSE}
plot(methyvim_out_rr, type = "volcano")
```

...

```{r methyvim-plot-pvals, eval=FALSE}
plot(methyvim_out_rr, type = "pvals")
```

...

```{r methyvim-plot-heatmap, eval=FALSE}
plot(methyvim_out_rr, type = "heatmap")
```

...

---

## References
